Novelty Detection Using Graphical Models for Semantic Room Classification

### 2011-05-16 semantic-nd/article.tex
Several changes on the results section.
That section should now be very close to final status.
Apart from the subsection titles.

% first the title is needed
\title{Using Graphical Models for Novelty Detection on Semantic Room Classification}

@@

\begin{abstract}
This papers presents an approach on implementing novelty detection for indoor room classification
using semantic data.
\emph{Graphical models} are used to model probabilistic knowledge and a novelty threshold is
defined in terms of conditional and unconditional probabilities.
The novelty threshold is then optimized using an unconditional probability density
model trained from unlabelled data.

@@

\begin{figure}[h]
\centering
\begin{tikzpicture}[node distance = 2cm, auto]
    \matrix[row sep=0.5cm,column sep=0.5cm] {
      & \node [var] (room) {Room Category}; \\
      \node [factor] (fAppearance) {}; &
      \node [factor] (fShape) {}; &
      \node [factor] (fSize) {}; \\

      \node [var] (appearance) {Appearance}; &
      \node [var] (shape) {Shape}; &
      \node [var] (size) {Size}; \\
    };
  \path [line] (room) -- (fAppearance);
  \path [line] (room) -- (fShape);
  \path [line] (room) -- (fSize);

  \path [line] (fAppearance) -- (appearance);
  \path [line] (fShape) -- (shape);
  \path [line] (fSize) -- (size);
\end{tikzpicture}
\caption{Example of a factor graph used for Dora for representing the environment.}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
% Introduce novelty detection
Novelty detection is an harder problem than that of classification due to its single-class nature.
Essentially a system is given training data and when given new data has to decide if it belongs
to the same concept it was trained with.

% Explain threshold approach to novelty detection
Due to the desire of robustness and generalization, a novelty detection system would always possess
some threshold that describes on how strict the system should be to new data.
This threshold is often seen as a distance measure to a normal input and several novelty detection
techniques are based on it.

@@

Using bayes rule on the reversed order $P(\overline{novel}|x)$ and assuming a constant $P(\overline{novel})$
we obtain a ratio between a conditional and unconditional probabilities of the input $x$.
Such a ratio is a suitable function for implementing an optimal novelty detector system with
thresholding.

@@

% Explain approach of using conditional probability
\subsection{Conditional Probability: $P(x|\overline{novel})$}
\label{sec:uniform-unconditional}

One of the hardest problems is that on most cases a novelty detection system only has access to 
the known data.
Under that the best approach is to define a threshold based on $P(x|\overline{novel})/P(x)$
considering that $P(x)$ is constant through all the input space.

In the presented case the conditional probability is approximated by the graphical model $G$
produced by the conceptual layer.
The \autoref{fig:conditional-prob-graph} represents a possible graph $G$ built from the conceptual
layer to represent the conditional probability on the sensed variables $x$.
$P_G(x)$ is used as an approximation for $P(x|\overline{novel})$.

@@

Its important to notice that in several cases assuming the unconditional probability $P(x)$
is constant leads to discarding the factor.
Nonetheless, here the distributions are dynamically changing as the system learns more on the environment.
So the normalizing argument $P(x)$ has to be evaluated for each new subset of $x$.

@@

Having now a graphical model built to model the known data distribution and assuming that in the unconditional case all
possible outcomes are equally likely a novelty threshold would be given by: $P_G(x)/P_U(x)$.

Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example the conditional probability would yield very small values on large sets of variables
$x$ than in small sets due to the spreading over the dimensions of the input space.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

\subsection{Using Unlabelled Data}
Nonetheless its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability
distribution than the uniform one assumed in \autoref{sec:uniform-unconditional}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
In order to verify the improvements obtained by using unlabelled data, a synthetic dataset
was generated.
The synthetic distribution assumes that an independent and variable size set of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 7 different measured properties types.

The known data was extracted by generating samples from 6 of the 11 classes.
Together with the generated known data, the room category that generated is given as label.
That known data is used to train factors to represent a conditional distribution.
Using unlabelled data, factors are trained to approximate an unconditional distribution
where all features are independent.
\autoref{fig:simple-experiment} shows the graph structure used for approximate the conditional and
unconditional distributions trained from the labelled and unlabelled data.

@@

\subfloat[Graph structure used to model conditional probability graph.
          The graph factors are trained using the labelled data.]{... conditional probability}
\qquad
\subfloat[Graph structure used to model unconditional probability graph.
          The graph factors are trained using the unlabelled data.]{... unconditional probability}

\caption{\label{fig:simple-experiment}The graph structures $G$ and $I$ used to model the
         conditional and unconditional probability.
         Note that graph $G$ correctly models the structure of the synthetic distribution, though
         during training it has not seen all possible room categories represented by variable $R_1$.}

\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.60\textwidth]{results/synthetic-all.pdf}

\caption{\label{fig:synthetic-roc}ROC curve comparing novelty detection performance between an optimal detector,
         $P_G(x)/P_U(x)$ a graph ratio considering an uniform unconditional distribution and
         $P_G(x)/P_I(x)$ a graph ratio considering an independent unconditional distribution}
\end{figure}

Since the distribution is synthetic its possible to create an optimal novelty detector, as
its the error rate $P(novel|x)$ can be exactly measured.
That serves as a comparison on how optimal a detector can be using the same input space.
A ROC curve was plotted for the three methods on \autoref{fig:synthentic-roc}.
As possible to see the unconditional probability plays an important role in obtaining a correct
ordering of the input space.

The importance of a correct estimate of unconditional probability is expected to decrease as the
input space increases and allows the several existing classes to become easily distinguished.

For comparison ROC curves where plot for the developed threshold based on the number of input features.

@@

\caption{\label{fig:synthetic-roc-breakdown}ROC curves plotted showing performance of the
         presented novelty detection method on graphs generated for different amount of
         sensed variables.
         As possible to see as the system gains more semantic information it becomes easier
         to detect novelty and the effect of picking an approximation for the unconditional
         probability decreases.}
\end{figure}

>>>>

% first the title is needed
\title{Novelty Detection Using Graphical Models for Semantic Room Classification}

@@

\begin{abstract}
This paper presents an approach to the problem of novelty
detection in the context of semantic room categorization.
The ability to assign semantic labels to areas in the environment is crucial for
autonomous agents aiming to perform complex human-like tasks and human
interaction.
However, in order to be robust and naturally learn the semantics from
the human user, the agent must be able to identify gaps in its own knowledge.
To this end, we propose a method based on graphical models to identify novel
input which does not match any of the previously learnt semantic descriptions.
The method employs a novelty threshold defined in terms of conditional
and unconditional probabilities. The novelty threshold is then optimized using
an unconditional probability density model trained from unlabelled data.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Novelty Detection}
% Introduce novelty detection
Novelty detection deals with detecting that a new input was generated by a class
other than the ones the system knows about.
It is harder than classification as only positive samples of a class are available
rendering normal classification methods invalid.

% Explain threshold approach to novelty detection
Due to the desire of robustness and generalization, a novelty detection system would
always possess some threshold that describes on how strict the system should be to
new data.
This threshold describes a trade of between generalization and certainty.
And is often seen as a distance measure to the known class.

@@

Using bayes rule on the reversed order $P(\overline{novel}|x)$ and assuming a constant $P(\overline{novel})$
a ratio between a conditional and unconditional probabilities of the input $x$ is obtained.
Such a ratio is a suitable function for implementing an optimal novelty detector system with
thresholding.

@@

% Explain approach of using conditional probability
\subsection{Conditional Probability}
\label{sec:conditional-prob}

In the presented case the conditional probability is approximated by the graphical model $G$
produced by the conceptual layer.
\autoref{fig:conditional-prob-graph} represents a possible graph $G$ built from the conceptual
layer to represent the conditional probability on the sensed variables $x$.
$P_G(x)$ is used as an approximation for $P(x|\overline{novel})$.

@@

\subsection{Unconditional Probability}
\label{sec:unconditional-prob}
One of the hardest problems is that on most cases a novelty detection system only
has access to the known data.
Under that the best approach is to define a threshold assuming that $P(x)$ is
constant through all the input space.

Its important to notice that in several cases assuming it to be constant leads to
discarding the factor.
Nonetheless, here the distributions are dynamically changing as the system learns
more on the environment.
So the normalizing argument $P(x)$ has to be evaluated for each new subset of $x$.

@@

Having a graphical model $G$ built to model the known data distribution and a model
$U$ for the unconditional probability a novelty threshold would be given by:
$P_G(x)/P_U(x)$.
Here $P_{U}(x)$ can be seen as a normalizing factor to lever all the $P_G(x)$ on any set of
variables $x$ into the same measure units (error rate), such that a static threshold can be implemented.
For example the conditional probability would yield very small values on large sets of variables
$x$ than in small sets due to the spreading over the dimensions of the input space.
As so a novelty measure is seen as a ratio on how much introducing the known concepts helps to
understand the observed result.

\subsection{Semi Supervised: Using Unlabelled Data}
Nonetheless its often the case that there is access to large amounts of unlabelled data.
Under that it becomes possible to obtain a better approach to the unconditional probability
distribution than the uniform one.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
In order to verify the performance of the proposed thresholds a synthetic dataset
was generated. To keep it simple only information regarding direct properties of
a room were modelled and no structured knowledge such as room connectivity was taken
in account.
The synthetic distribution assumes that an independent and variable size set of features
$x$ is generated by a given room category.
In whole there was 11 different room categories and 7 different measured feature
types. Each feature type can be present more than once in (Eg.: room shape is
extracted from 2D laser scans in more than one position in the room).

The room categories were chosen to mimic as close as possible real features and
categories existent on reality. And they model different room categories with
different levels of detail. For example: 1 person office, 2 person office, hallway,
robot lab.

From the distribution, 100 labelled samples for 5 of the 11 room categories were
drawn to represent the known concepts and 1000 unlabelled samples were drawn from
all the room categories for learning the unconditional probability distribution.
Using those samples, factors were learnt for the graphs used to model the
conditional distribution and the independent unconditional distribution.
\autoref{fig:simple-experiment} shows the graph structure used for approximate the
trained conditional and unconditional distributions.
Its important to notice that graph $G$ used to model the known classes perfectly
mimics the real distribution, with the exception that the system is not aware of
all the room categories.

@@

\subfloat[Graph structure $G$ used to model conditional probability graph.]{... conditional probability}
\qquad
\subfloat[Graph structure $U$ used to model unconditional probability graph.]{... unconditional probability}
\qquad
\subfloat[Graph structure $I$ used to model unconditional probability graph.]{... unconditional probability}

\caption{\label{fig:simple-experiment}The graph structures used to model the
         conditional and unconditional probability for implementing the novelty
         thresholds $P_G(x)/P_U(x)$ and $P_G(x)/P_I(x)$.}
\end{figure}

% Describe how to obtain the 3 thresholds functions seen on the graphs.
Using the learned models $G$, $U$ and $I$ two thresholds were trained:
$P_G(x)/P_U(x)$ assuming a uniform unconditional distribution
and $P_G(x)/P_I(x)$ assuming an independent unconditional distribution.
Since the distribution is synthetic there is access to $P(x)$ and $P(x|concept)$
and so a perfect threshold function could also be created to test how far the
presented thresholds are from optimal.

\subsection{Suitable functions for implementing a Static Threshold}
%%% Results 1
% Show the threshold ratio is an optimal detector (if perfect information was available)
% Show that the thresholds are suitable functions for implementing a static threshold.
As first tryout, the performance of the novelty thresholds was plotted for a set
of 1000 samples taken out from the whole distribution (\autoref{fig:synthetic-roc}).
Those samples contain different number of sensed properties for each room, miming
the dynamic properties expected to see when implemented on a robot.

\begin{figure}[h]
\centering
\includegraphics[width=0.60\textwidth]{results/synthetic-all.pdf}

\caption{\label{fig:synthetic-roc}ROC curve comparing novelty detection performance
         under samples with variable size of sensed properties.}
\end{figure}

The convex shape of the optimal threshold shows that the ratio between conditional
and unconditional probability is indeed an optimal detector and is suitable for
implementing a static threshold when the samples are taken from dynamic
distributions (eg.: some samples where there is only access to room size versus
samples where there is a lot of information on the room properties).

% Discuss importance on approximating unconditional probability.
Its also possible to see how important it is to estimating a correct unconditional
probability in order to obtain a correct novelty ordering on the inputs.
The assumption of an uniform unconditional probability has led to very poor results.
That is probably explained by semantic properties as the ones modelled to be highly
biased towards some values and understanding that bias plays an important step
to detect whether a given sensed value is a valuable clue on the room category
or not.

%%% Results 2
% Measure performance of the thresholds as more information becomes available.
\subsection{Performance impact as more information becomes available}
In order to measure the performance impact as more semantic information becomes
available ROC curves were plotted for samples grouped by the number of sensed
semantic properties.

@@

\caption{\label{fig:synthetic-roc-breakdown}ROC curves plotted showing performance of the
         presented novelty detection method on graphs generated for different amount of
         sensed variables.}
\end{figure}

As possible to see as the system gains more semantic information it becomes easier
to detect novelty as the input space size increases and allows the several existing
classes to become more easily distinguished.

The performance of the independent threshold decreases as the number of sensed
variables increases. This is easily explained by the fact that the graph $I$ is not
able to model the existent dependence between the variables and that becomes obvious
as the number of variables increases (Eg.: graph $I$ perfectly models $P(x)$ in the
case where only 1 variable is present).

The uniform threshold shows a very poor performance specially on small size samples
where its performs almost no better than random.
It performance increases as the size of sensed variables increases but nonetheless
its always very small when compared to how optimal a threshold could be.
This shows that at least under semantic distributions as the one modelled with the
synthetic data assuming an uniform unconditional distribution is a very bad choice.
